---
title: "tidyr"
author: "Robert J. Gamble"
date: "October 23, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse) # install.packages("tidyverse")

## read in gapminder data in wide format
gap_wide <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv')

```


## tidyr gather()
```{r}
gap_long <- gap_wide %>% 
  gather(key = obstype_year, value = obs_values, 
         -continent, -country)


```


## tidyr separate()
```{r}
gap_long <- gap_wide %>% 
  gather(key = obstype_year, value = obs_values, 
         -continent, -country) %>% 
  separate(col = obstype_year, 
           into = c("obstype", "year"), 
           sep = "_",
           convert = TRUE) # converts to integer(otherwise year would be characters)


```

```{r}
canada_df <- gap_long %>% 
  filter(country == "Canada", obstype == "lifeExp")

ggplot(data = canada_df, aes(x = year, y = obs_values)) +
  geom_line()
```

Plot life expectancy of all countries in the Americas

```{r}
life_df <- gap_long %>% 
  filter(obstype == "lifeExp", continent == "Americas")
ggplot(data = life_df, aes(x = year, y = obs_values, color= country)) +
  geom_line()


```

Exercise: 

Using gap_long, calculate and plot the the mean life expectancy for each continent over time from 1982 to 2007. Give your plot a title and assign x and y labels. Hint: use the dplyr::group_by() and dplyr::summarize() functions.

```{r}

lifePerCont_df <- gap_long %>% 
  filter(obstype == "lifeExp", year > 1980) %>% 
  group_by(continent, year) %>%
  summarize(mean_LifeExp = mean(obs_values)) %>%
  ungroup()

ggplot(data = lifePerCont_df, aes(x = year, y = mean_LifeExp, color= continent)) +
  geom_line() +
  labs( x = "Year",
        y = "Age (years)",
        title = "Mean life expectancy") +
  theme_bw()


```

# spread()
```{r}

gap_normal <- gap_long %>%  
  spread(key = obstype,
         value = obs_values)
```

Exercise:

Convert “gap_long” all the way back to gap_wide. Hint: you’ll need to create appropriate labels for all our new variables (time*metric combinations) with the opposite of separate: tidyr::unite().

```{r}

gap_wide2 <- gap_long %>% 
  unite(col = var_names, obstype, year, sep = "_") %>% 
  spread(var_names, obs_values)
```

